### 2021-04-28 paper/parlr_paper_1.tex
finished big simulation. updated parlr paper

@@

Most record linkage techniques are derived from the seminal 1969 paper
by Fellegi and Sunter: ``A Theory for Record Linkage''. The defining
characteristic of their model was to transform the sets of records,
which often contain text data that is difficult to model, into sets of
comparison vectors governed by parameters that can be more easily
estimated. Concretely, if files \(A\) and \(B\) have \(n_A\) and \(n_B\)
records respectively, and if the files share \(F\) fields in common upon
which to base the linkage, the Fellegi and Sunter approach generates a
\(n_A n_B \times F\) matrix \(\Gamma\), which contains similarity scores
between each pair of records across datasets. We say \(\gamma_{ij}\) is
the comparison vector for record \(i \in A\) and record \(j \in B\),
with \(\gamma_{ij}^f\) providing their similarity score on the
\(f^{th}\) field. For ease of modeling and computation, we restrict
these similarity scores to be discrete, ordinal variables, and the
construction of these is left to the modeller. It is common to use use
binary 0-1 variables to indicate exact matching, and 0-1-2 variables to
provide an option for partial matching. For text data, we calculate
similarity based on Levenstein distance or some other text similarity
score, and bin these scores to integers for use in the model.

Two modern adaptations of the Fellegi Sunter model are important for
understanding \texttt{parlr} and its contribution. In 2019, Enamorado et
al proposed \texttt{fastlink}, a method that closely followed the
modelling assumptions of Fellegi and Sunter, but used innovating hashing
techniques to greatly enhance its speed and scale. Importantly, they
maintained Fellegi and Sunter's independent matching assumption, in
which the matching status of any \((i,j)\) pair is made independently of
the matching of any other \((i', j')\). This leads to many matchings
that violate ``one-to-one'' considerations that are often implicit in
the data, which need to be resolved in a post-processing step. In
contrast, Sadinle (2017) proposed ``Beta Record Linkage for Bipartite
Matching'' and the accompanying \texttt{BRL} package which strictly
enforces one-to-one matching. Specifically, in each iteration of his
Gibbs sampler, he considers each record \(j\in B\), removes from
consideration the records \(i\in A\) that have already been matched,
then samples a potential link. However, accounting for these
dependencies throughout the linkage process is computionally burdensome,
leaving \texttt{BRL} only suitable for small to moderate linkage
problems.

Outside of the Fellegi-Sunter framework, some researchers have developed
methods that model the data directly, not a derived set of comparison
vectors. As an early example, Steorts et al 2016 produced
\texttt{blink}, which had the advantage of a computational complexity
that grew linearly with the data (as opposed to quadratically within
Fellegi Sunter), but was plagued with slow mixing times and was unable
to incorportate text data. Later, Steorts et al 2020 improved on this
method with \texttt{dblink}, which was able to simulate text data within
its Gibbs sampler by drawing from an empirical distribution, and
remedied its slow mixing through a probabalistic blocking approach.
However, this method remains computationally intensive and requires
distributed computing to manage even moderate linkage problems in
acceptable time.

@@

@@

Our Gibbs sampler provides posterior samples of \(\mathbf{Z}\) which we
use to make our final decisions about the linkage structure. In the case
where we false matches and missed matches contribute the same loss, we
declare \((i,j)\) to be a match whenever \(P(Z_j = i) > \frac{1}{2}\)
according to these posterior samples. We can create more elaborate
decisions by attributing different loss values to different kinds of
error, and allowing pairings with middling posterior probabilities to be
left unlabeled by the algorithm so that the modeler can address those
manually. For further discussion, see Sadinle 2017.

Since our Gibbs procedure does not strictly enforce one-to-one matching,
it is possible for the final Bayes estimate to link mulitple records in
\(B\) to one record in \(A\). The modeller can either report both such
matches (with their respective posterior match probabilities), or
resolve these conflicts by accepting only the match with highest
posterior probability. Rigorous proof of this resolution procedure is
deferred to future work.

@@

To address this, we propose a new strategy called \emph{random
indexing}. In standard indexing, the modeller decides a certain criteria
that they expect all true matching pairs to satisfy, and a priori label
any record pairs that do not meet that criteria as non-matching. For
example, one might only consider pairs with a certain similarity score
on a field deemed to be important (like first name), or only pairs with
exact matching on a specified number of fields. While generally chosen
to be be quite loose, establishing these criteria requires knowledge of
the problem and invites room for human error. We propose a method of
reducing the comparison space (and reducing storage requirements)
without these drawbacks. Note that all records of the same agreement
pattern have the same probality when sampling \(Z_j\). Therefore we know
that records belonging to an \(h_p\) such that \(H_{p_j}\) is large are
very unlikely to be sampled consistently enough to be deemed a match
through the Bayes estimate, even without considering the form of the
agreement pattern itself. In random indexing, rather than store all of
these unlikely record labels, we choose to store only a small number
\(R\) of them. Posterior calculations still attribute the appropiate
weight to all records through the summary statistics \(H_{p_j}, H_p^m\),
and \(H_p^u\); this is similar to the \emph{u-correction} concept
introduced by \_\_\_\_\_\_\_\_. Rather than storing \(n_A \times n_B\)
record labels, random indexing allows us to store at most
\(n_B \times P \times R\), regardless of how large \(n_A\) is. Coupled
with the partitioning strategy for creating comparison vectors, this
greatly expands the size of the linkage tasks we can undertake.

@@

After receiving matching statuses from \(\mathbf{Z}\), the Sadinle
method calculates \(\alpha_{fl}(\mathbf{Z})\) and
\(\beta_{fl}(\mathbf{Z})\) for each field and level. This constitutes
\(2 \times \sum L_f\) many summations over \(n_A \times n_B\)
quantities, and becomes computationally burdensome with large data. In
contrast, we recognize that each unique agreement pattern contributes to
the posterior \(\alpha(\mathbf{Z})\) and \(\beta(\mathbf{Z})\) vectors
in the same way. In fact, if we denote
\(H_p^m = \sum_{j=1}^{n_B} \mathbf{1}_{(Z_j, j) \in h_p}\) to be the
number of matching record pairs with agreement pattern \(p\), then the
contribution of pairs of pattern \(p\) to the \(\alpha(\mathbf{Z})\)
vector is simply \(H_p^m \times h_p\). Thus our posterior update for
\(\alpha\) is simply
\(\alpha(\mathbf{Z}) = \alpha_0 + \sum_{p=1}^P H_p^m \times h_p\). Then,
we can easily calculate \(H_p^u\), the number of nonmatching record
pairs of agreement pattern \(p\), by subtracting the number of matching
pairs from the total present in the data; that is
\(H_p^u = H_p - H_p^m\). From this, we can update our \(\beta\)
parameter through
\(\beta(\mathbf{Z}) = \beta_0 + \sum_{p=1}^P H_p^u \times h_p\). Note
that these constitute \(P\) many summations over \(n_B\) quantities, and
thus avoid the \(n_A \times n_B\) summation from the original method.

Sadinle uses a prior for \(\mathbf{Z}\) that induces the a full
conditional for \(Z_j\) that strictly enforces one-to-one matching. In
particular, this sampler removes previously matches records from the set
of candidate records when sampling \(Z_j\), creating a dependency that
makes the sampler \emph{inherently serial}. By weakening the one-to-one
requirement, our full conditional for \(Z\) does not depend on the rest
of the \(\mathbf{Z}_{-j}\) vector, and thus can be computed in parallel.
More importantly, since only the agreement pattern of \(Z_j\) is used
for calculations within the Gibbs sampler, and not the particular record
label, we can conduct this sampling only at the level of the unique
agreement patterns. This boosts computation time far greater than
parallelization.

To do this, we calculate the Fellegi Sunter weight \(w_{h_p}\) for each
unique pattern, sample the agreement pattern between \(j\) and its
potential match, and then sample the record label uniformly among viable
records. More concretely, define \(h(Z_j)\) to be the agreement pattern
between \(j\) and its potential match, and say \(h(Z_j) = h_{P+1}\) when
\(Z_j = n_A + 1\). Then,

@@

We complete the entire Gibbs procedure at the level of the \(P\)
agreement patterns. After, we can back-fill the records corresponding to
the agreement patterns by sampling uniformly at random among candidate
records stored in \(\tilde{\Gamma}\).

@@

To demonstrate speed, we generate comparison vectors from pre-specified
distributions so that we can easily increase the size of the linkage
problem. Distributions are meant to emulate the behavior of similarity
scores across first name, last name, and day, month, and and year of
birth, with 1 indicating an exact match on a particular field and 0
indicating a nonmatch. We simulate these data for different values of
\(n_A\) and \(n_B\), and compare the runtime of \texttt{parlr} against
\texttt{BRL}. We see that at low data size, \texttt{BRL} outperforms,
but that \texttt{parlr} is significantly faster at handling larger data.

In particular, runtime for \texttt{BRL} seems to grow quadratically
while runtime for \texttt{parlr} seems to grow linearly.

@@

The above discussion suggests that for fixed \(n_B\), computation time
should remain mostly constant with growing \(n_A\). Shockingly, this
seems to be true. In the plot below, fixing \(n_B = 500\), we see linear
growth for the runtime under \texttt{BRL} as \(n_A\) increases, with
much more static runtime under \texttt{parlr}. The slight increases in
runtime that we do see are due primarily to the hashing step, which
again can be run in parallel for large data.

@@

We note here that \texttt{BRL} is coded in C, which makes for unfair
comparison against \texttt{parlr}, currently only built in R.
Additionally, although \texttt{parlr} is amenable to parallezation, this
simulation was run on a single core. Running \texttt{parlr} in C++ with
paralellization for the hashing step and sampling the matching status of
the record pairs should lead to even more drastic results.

@@

\begin{itemize}
\tightlist
\item
  Necessary? El Salvador data? Or perhaps needs to be a bigger dataset?
\end{itemize}

@@

We note that this often not a reasonable assumption; for example if we
know that two records differ on gender, this does not give any
information on whether their year of births are the same, but it does
suggest that its more likely that their first names are different.

>>>>

\hypertarget{introduction-or-abstract}{%
\subsection{Introduction (or
Abstract?)}\label{introduction-or-abstract}}

Record linkage is the task of identifying duplicate records across
multiple data sources. This task is straightforward when provided unique
identifiers or highly reliable information, but becomes difficult when
data sources are plagued with error. Additionally, record linkage is
inherently computationally intensive, and many existing methods are not
feasible for many applications. In this paper, we propose a Bayesian
model for record linkage that is amenable to parallel computing, hashing
techniques and efficient storage mechanisms, that is scalable to large
linkage tasks.

@@

Most record linkage techniques are derived from the seminal 1969 paper
by Fellegi and Sunter: ``A Theory for Record Linkage''. The defining
characteristic of their model was to transform the sets of records,
which often contain text data that is difficult to model, into sets of
comparison vectors governed by parameters that can be more easily
estimated. Concretely, if files \(A\) and \(B\) have \(n_A\) and \(n_B\)
records respectively, and if the files share \(F\) fields in common upon
which to base the linkage, the Fellegi and Sunter approach generates a
\(n_A n_B \times F\) matrix \(\Gamma\), which contains similarity scores
between each pair of records across datasets. We say \(\gamma_{ij}\) is
the comparison vector for record \(i \in A\) and record \(j \in B\),
with \(\gamma_{ij}^f\) providing their similarity score on the
\(f^{th}\) field. For ease of modeling and computation, we restrict
these similarity scores to be discrete, ordinal variables, and the
construction of these is left to the modeler. It is common to use use
binary 0-1 variables to indicate exact matching, and 0-1-2 variables to
provide an option for partial matching. For text data, we calculate
similarity based on Levenstein distance or some other text similarity
score, and bin these scores to integers for use in the model.

Two modern adaptations of the Fellegi Sunter model are important for
understanding \texttt{parlr} and its contribution. In 2019, Enamorado et
al proposed \texttt{fastlink}, a method that closely followed the
modelling assumptions of Fellegi and Sunter, but used innovating hashing
techniques to greatly enhance its speed and scale. Importantly, they
maintained Fellegi and Sunter's independent matching assumption, in
which the matching status of any \((i,j)\) pair is made independently of
the matching of any other \((i', j')\). This leads to many matchings
that violate ``one-to-one'' considerations that are often implicit in
the data, which need to be resolved in a post-processing step. In
contrast, Sadinle (2017) proposed ``Beta Record Linkage for Bipartite
Matching'' and the accompanying \texttt{BRL} package which strictly
enforces one-to-one matching. Specifically, in each iteration of his
Gibbs sampler, he considers each record \(j\in B\), removes from
consideration the records \(i\in A\) that have already been matched,
then samples a potential link. However, accounting for these
dependencies throughout the linkage process is computationally
burdensome, leaving \texttt{BRL} only suitable for small to moderate
linkage problems.

Outside of the Fellegi-Sunter framework, some researchers have developed
methods that model the data directly, not a derived set of comparison
vectors. As an early example, Steorts et al 2016 produced
\texttt{blink}, which had the advantage of a computational complexity
that grew linearly with the data (as opposed to quadratically within
Fellegi Sunter), but was plagued with slow mixing times and was unable
to incorporate text data. Later, Steorts et al 2020 improved on this
method with \texttt{dblink}, which was able to simulate text data within
its Gibbs sampler by drawing from an empirical distribution, and
remedied its slow mixing through a probabilistic blocking approach.
However, this method remains computationally intensive and requires
distributed computing to manage even moderate linkage problems in
acceptable time.

@@

Here the reader should the relationship between our proposed model and
that of Sadinle 2017. In his model, Sadinle constructs a prior
distribution on the entire \(\mathbf{Z}\) vector, which induces a Gibbs
sampler that strictly enforces one-to-one matching. In particular, this
sampler removes previously matches records from the set of candidate
records when sampling \(Z_j\), creating a dependency that makes the
sampler \emph{inherently serial}. We however use independent priors for
each \(Z_j\), thereby weakening the one-to-one requirement; our sampler
does ensure that each record in \(B\) can be matched to at most one
record in \(A\), but allows for the possiblity that multiple records in
\(B\) match to the same record in \(A\). With this change, components of
the \(\mathbf{Z}\) vector can be computed in parallel, allowing for
significant computational gains. More importantly, since only the
agreement pattern of \(Z_j\) is used for calculations within the Gibbs
sampler, and not the particular record label, we can conduct this
sampling only at the level of the unique agreement patterns, offering
even more computational savings. This will be explained further in
\textcolor{red}{Section XX}.

@@

(Adapt from Sadinle)

Since our Gibbs procedure does not strictly enforce one-to-one matching,
it is possible for the final Bayes estimate to link multiple records in
\(B\) to one record in \(A\). The modeler can either report both such
matches (with their respective posterior match probabilities), or
resolve these conflicts by accepting only the match with highest
posterior probability. Rigorous proof of this resolution procedure is
deferred to future work.

@@

To address this, we propose a new strategy called \emph{storage
efficient indexing} (SEI). In standard indexing, the modeler decides a
certain criteria that they expect all true matching pairs to satisfy,
and a priori label any record pairs that do not meet that criteria as
non-matching. For example, one might only consider pairs with a certain
similarity score on a field deemed to be important (like first name), or
only pairs with exact matching on a specified number of fields. While
generally chosen to be be quite loose, establishing these criteria
requires knowledge of the problem and invites room for human error. We
propose a method of reducing the comparison space (and reducing storage
requirements) without these drawbacks. Note that all records of the same
agreement pattern have the same probability when sampling \(Z_j\).
Therefore we know that records belonging to an \(h_p\) such that
\(H_{p_j}\) is large are very unlikely to be sampled consistently enough
to be deemed a match through the Bayes estimate, even without
considering the form of the agreement pattern itself. In random
indexing, rather than store all of these unlikely record labels, we
choose to store only a small number \(R\) of them. Posterior
calculations still attribute the appropriate weight to all records
through the summary statistics \(H_{p_j}, H_p^m\), and \(H_p^u\); this
is similar to the \emph{u-correction} concept introduced by
\_\_\_\_\_\_\_\_. Rather than storing \(n_A \times n_B\) record labels,
random indexing allows us to store at most \(n_B \times P \times R\),
regardless of how large \(n_A\) is. Coupled with the partitioning
strategy for creating comparison vectors, this greatly expands the size
of the linkage tasks we can undertake.

@@

\underline{Updating $\mathbf{m}$ and $\mathbf{u}$:} After receiving
matching statuses from \(\mathbf{Z}\), the Sadinle method calculates
\(\alpha_{fl}(\mathbf{Z})\) and \(\beta_{fl}(\mathbf{Z})\) for each
field and level. This constitutes \(2 \times \sum L_f\) many summations
over \(n_A \times n_B\) quantities, and becomes computationally
burdensome with large data. In contrast, we recognize that each unique
agreement pattern contributes to the posterior \(\alpha(\mathbf{Z})\)
and \(\beta(\mathbf{Z})\) vectors in the same way. In fact, if we denote
\(H_p^m = \sum_{j=1}^{n_B} \mathbf{1}_{(Z_j, j) \in h_p}\) to be the
number of matching record pairs with agreement pattern \(p\), then the
contribution of pairs of pattern \(p\) to the \(\alpha(\mathbf{Z})\)
vector is simply \(H_p^m \times h_p\). Thus our posterior update for
\(\alpha\) is simply
\(\alpha(\mathbf{Z}) = \alpha_0 + \sum_{p=1}^P H_p^m \times h_p\). Then,
we can easily calculate \(H_p^u\), the number of nonmatching record
pairs of agreement pattern \(p\), by subtracting the number of matching
pairs from the total present in the data; that is
\(H_p^u = H_p - H_p^m\). From this, we can update our \(\beta\)
parameter through
\(\beta(\mathbf{Z}) = \beta_0 + \sum_{p=1}^P H_p^u \times h_p\). Note
that these constitute \(P\) many summations over \(n_B\) quantities, and
thus avoid the \(n_A \times n_B\) summation from the original method.

\underline{Updating $\mathbf{Z}$:} Although sampling \(Z_j\) from a the
full conditional provided earlier is conceptually straightforward, it
becomes computational burdensome when \(n_A\) is larger. The reader can
confirm that sampling a value from a large set of unequal probabilities
becomes difficult in most programming languages. To speed up
computation, we break this sampling step into two simpler steps. First,
we calculate the Fellegi Sunter weight \(w_{h_p}\) associated with each
unique pattern and sample the agreement pattern between \(j\) and its
potential match. Second, we sample the record label uniformly among
records associated with that agreement pattern for that particular
\(j\in B\). More concretely, define \(h(Z_j)\) to be the agreement
pattern between \(j\) and its potential match, and say
\(h(Z_j) = h_{P+1}\) when \(Z_j = n_A + 1\). Then,

@@

Lastly, we recognize that all posterior updates are governed by the
agreement patterns of the record pairs rather than the record labels
themselves. Thuse We complete the entire Gibbs procedure at the level of
the \(P\) agreement patterns. After, we can back-fill the records
corresponding to the agreement patterns by sampling uniformly at random
among candidate records stored in \(\tilde{\Gamma}\). Sampling uniformly
is computational simple even for large sets of candidate records, but
this step can also be parallelized when working with large data.

@@

To demonstrate speed, we generate comparison vectors from pre-specified
distributions so that we can easily increase the size of the linkage
problem. Distributions are meant to emulate the behavior of similarity
scores across first name, last name, and day, month, and and year of
birth, with 1 indicating an exact match on a particular field and 0
indicating a nonmatch. We simulate these data for different values of
\(n_A\) and \(n_B\), and compare the run-time of \texttt{parlr} against
\texttt{BRL}. We see that at low data size, \texttt{BRL} outperforms,
but that \texttt{parlr} is significantly faster at handling larger data.

In particular, run-time for \texttt{BRL} seems to grow quadratically
while run-time for \texttt{parlr} seems to grow linearly.

@@

The above discussion suggests that for fixed \(n_B\), computation time
should remain mostly constant with growing \(n_A\). Shockingly, this
seems to be true. In the plot below, fixing \(n_B = 500\), we see linear
growth for the run-time under \texttt{BRL} as \(n_A\) increases, with
much more static run-time under \texttt{parlr}. The slight increases in
run-time that we do see are due primarily to the hashing step, which
again can be run in parallel for large data.

@@

We note here that \texttt{BRL} is coded in C, which makes for unfair
comparison against \texttt{parlr}, currently only built in R.
Additionally, although \texttt{parlr} is amenable to parallelization,
this simulation was run on a single core. Running \texttt{parlr} in C++
with paralellization for the hashing step and sampling the matching
status of the record pairs should lead to even more drastic results.

@@

\begin{itemize}
\tightlist
\item
  Thinking of re-running the El Salvador linkage from the Sadinle 2017
  paper, but doing so with without the formal blocking and indexing that
  he uses. Instead, I'd use hashing and storage efficient indexing -- I
  should be able to do linkage on the all the data at once!
\end{itemize}

@@
